Extracting production locations for each species to obtain environmental forcings

### 2025-07-09 01_tidying_data.qmd
Working with rdsi data

```{r globals}
# Specify species of interest
species <- c(
  "atlantic_salmon",
  "general_salmonid",
  "european_seabass",
  "gilthead_seabream",
  "large_yellow_croaker",
  "japanese_seabass",
  "japanese_amberjack",
  "general_grouper",
  "pompano",
  "milkfish"
)

@@

#rename the NASA files

#SST_files <- list.files(file.path(rdsi_dir, "raw_data/NASA/GHRSST_L4_MUR_0.25_SST"), full.names = FALSE, pattern = ".nc")

# map(SST_files, \(this_file){
#   file.rename(from = file.path(nasa_dir, this_file), to = file.path(nasa_dir, sub('^(.{4})(.*)$', '\\1_\\2',  this_file)))
# })

@@

# this_file <- SST_files[[1]]
# 
# this_year=2013

#initialise lists for each year

years <- c(2010:2019)
year_lists <- vector(mode ="list", length = length(years))

# Add annual files to a list of files for each year

for(y in 1:length(years)){

  this_year <- years[y]

  this_years_files <- SST_files[grep(paste0(this_year, "_"), SST_files)]

  this_years_files <- this_years_files[!grepl(paste0(this_year, "_0229"), this_years_files)]

  year_lists[[y]] <- this_years_files

}

# Extract the daily files for each year and rasterize in a stack

days <- c(1:365)

this_day <- days[1]

rast_list = vector(mode = "list", length = length(days))

map(days, \(this_day){

  this_days_file_list <- sapply(year_lists, "[[", this_day)

  message("SST layer for Day ", this_day)

  this_day_stack <- rast(this_days_file_list, subds = "analysed_sst")

  this_day_mean_rast <- app(this_day_stack, "mean", na.rm=TRUE)-273.15

  writeRaster(this_day_mean_rast, filename = sprintf(here("data/_general_data/SST/SST_rasters/sst_nasa_mur_L4_0.25_mean2010-2019_day_%s.tif"), this_day))

})

# now gapfill the saved rasters

raw_SST_files <- list.files("data/_general_data/SST/SST_rasters/", full.names = TRUE)

@@

```{r}

vuln_scores <- read.csv(file.path(rdsi_data_dir, "marine_spp_vulnerabilities/butt_spp_vuln_framework-publication/_output/vuln_gapfilled_score.csv")) %>% 
  select(vuln_gf_id, eutrophication_nutrient_pollution)

@@

qsave(x = full_vuln_df, file = here("data/_general_data/species_layers/vulnerabilities/marine_spp_vulnerabilities_eutrophication.qs"))

```

@@

#save to project
qsave(x = aquamaps_0.5d, file = here("data/_general_data/species_layers/distribution/aquamaps_0.5d.qs"))

```

>>>>

```{r globals}
# Specify species of interest
species <- c("atlantic_salmon", "general_salmonid")

@@

#rename the NASA files 
#SST_files <- list.files(file.path(rdsi_dir, "raw_data/NASA/GHRSST_L4_MUR_0.25_SST"), full.names = FALSE, pattern = ".nc")
# map(SST_files, \(this_file){
#   file.rename(from = file.path(nasa_dir, this_file), to = file.path(nasa_dir, sub('^(.{4})(.*)$', '\\1_\\2',  this_file)))
# })

@@

# this_file <- SST_files[[1]]
# this_year=2013

#initialise lists for each year
years <- c(2010:2019)
year_lists <- vector(mode ="list", length = length(years))

# Add annual files to a list of files for each year
for(y in 1:length(years)){
  this_year <- years[y]
  this_years_files <- SST_files[grep(paste0(this_year, "_"), SST_files)]
  this_years_files <- this_years_files[!grepl(paste0(this_year, "_0229"), this_years_files)]
  year_lists[[y]] <- this_years_files
}

# Extract the daily files for each year and rasterize in a stack
days <- c(1:365)
this_day <- days[1]
rast_list = vector(mode = "list", length = length(days))

map(days, \(this_day){
  this_days_file_list <- sapply(year_lists, "[[", this_day)
  message("SST layer for Day ", this_day)
  this_day_stack <- rast(this_days_file_list, subds = "analysed_sst")
  this_day_mean_rast <- app(this_day_stack, "mean", na.rm=TRUE)-273.15
  writeRaster(this_day_mean_rast, filename = sprintf(here("data/_general_data/SST/SST_rasters/sst_nasa_mur_L4_0.25_mean2010-2019_day_%s.tif"), this_day))
})

# now gapfill the saved rasters
raw_SST_files <- list.files("data/_general_data/SST/SST_rasters/", full.names = TRUE)

@@

```{r}
vuln_scores <- read.csv(file.path(rdsi_data_dir, "marine_spp_vulnerabilities/butt_spp_vuln_framework-publication/_output/vuln_gapfilled_score.csv")) %>% 
  select(vuln_gf_id, eutrophication_nutrient_pollution)

@@

qsave(x = full_vuln_df, file = here("data/_general_data/species_layers/vulnerabilities/marine_spp_vulnerabilities_eutrophication.qs"))
```

@@

#save to project
qsave(x = aquamaps_0.5d, file = here("data/_general_data/species_layers/distribution/aquamaps_0.5d.qs"))
```

### 2025-07-09 02_extracting_production_locations.qmd
Working with rdsi data

```{r setup}
library(tidyr)
library(dplyr)
library(magrittr)
library(here)
library(janitor)
library(sf)
library(qs)
library(countrycode)
library(terra)
library(units)

@@

Isolate production data for Atlantic salmon.
```{r rich-raw-processing}
#| eval: false

@@

(mariculture_spp_prod <- 
    aquaculture_raw %>% 
    clean_names() %>% 
    select(c(1:7, starts_with("x"))) %>% 
    pivot_longer(names_to = "year", values_to = "production", cols= -c(1:7)) %>% 
    mutate(year = gsub(pattern="x", replacement = "", year)) %>% 
    filter(year %in% c(2017)) %>% 
    filter(!grepl("Inland waters", fao_major_fishing_area_name)  & 
             !asfis_species_name_3 %in% c("Clams, cockles, arkshells", 
                                          "Abalones, winkles, conchs",
                                          "Mussels", 
                                          "Oysters",
                                          "Shrimps, prawns",
                                          "Crabs, sea-spiders",
                                          "Freshwater molluscs",
                                          "Squids, cuttlefishes, octopuses",
                                          "Pearls, mother-of-pearl, shells",
                                          "Miscellaneous aquatic plants",
                                          "Scallops, pectens"),
           !grepl("seaweeds" , asfis_species_name_3),
           !grepl("molluscs|cucumber|invertebrates", asfis_species_name_2)) %>% 
    group_by(country_name, asfis_species_name_2, asfis_species_name_3, fao_major_fishing_area_name, fao_major_fishing_area_code, year) %>% 
    summarise(production = sum(production)) %>% 
    ungroup() %>% 
    mutate(asfis_species_name_2 = case_when(asfis_species_name_2 == "Marine fishes nei" & country_name == "Viet Nam" ~ "Groupers nei",
                                            TRUE ~ asfis_species_name_2)) %>% 
    mutate(iso3c = countrycode(sourcevar = country_name, origin = "country.name", destination = "iso3c", warn =TRUE)) %>% 
    group_by(iso3c, country_name, year) %>% 
    nest() %>% 
    mutate(national_production = map(data, ~(sum(.$production)))) %>% 
    unnest(cols = c(data, national_production)) %>% 
    ungroup() %>% 
    filter(asfis_species_name_2 %in% spp_names) %>% 
    mutate(prop_production = production/national_production) %>% 
    filter(!asfis_species_name_2 %in% c("Marine fishes nei")) %>% 
  mutate(model_name = case_when(grepl("rouper", asfis_species_name_2) ~ "general_grouper", 
                                asfis_species_name_2 %in% c("Coho(=Silver) salmon", "Rainbow trout") ~ "general_salmonid",
                                asfis_species_name_2 %in% c("Amberjacks nei") ~ "japanese_amberjack",
                                TRUE ~ gsub(pattern = " ", replacement = "_", tolower(asfis_species_name_2)))) %>% 
  relocate(model_name, .before = asfis_species_name_2) %>% 
  filter(prop_production>0) %>% 
    mutate(farm_name = case_when(model_name %in% c("atlantic_salmon", "general_salmonid") ~ "Salmonidae fish",
                                 TRUE ~ "General marine fish"))
    )

>>>>

```{r setup}
library(tidyr)
library(dplyr)
library(magrittr)
library(here)
library(janitor)
library(sf)
library(qs)
library(countrycode)
library(terra)
library(units)
library(readr)

@@

Clean aquaculture production data from the FAO and isolate production data for Atlantic salmon.

```{r clean-raw-data}
# Get aquaculture quantity in tonnes live weight
quantity <- file.path(rawdata_path, "Aquaculture_2025.1.0", "Aquaculture_Quantity.csv") %>% 
  read.csv() %>% 
  clean_names() %>% 
  mutate(value = set_units(value, "t")) %>% 
  select(-measure)

# Apply environment codes and filter by "marine"
environment_codes <- file.path(rawdata_path, "Aquaculture_2025.1.0", "CL_FI_PRODENVIRONMENT.csv") %>% 
  read.csv() %>% 
  distinct(Code, Name_En)

quantity <- quantity %>% 
  merge(environment_codes, by.x = "environment_alpha_2_code", by.y = "Code") %>%
  rename(environment = Name_En) %>% 
  filter(environment == "Marine") %>% 
  select(-environment_alpha_2_code)

# Apply species codes and filter by relevant species groups
species_codes <- file.path(rawdata_path, "Aquaculture_2025.1.0", "CL_FI_SPECIES_GROUPS.csv") %>% 
  read.csv() %>% 
  filter(str_detect(Scientific_Name, "Salmo")) %>% 
  filter(Scientific_Name %in% c("Salmoniformes (=Salmonoidei)", "Salmo spp", "Salmo salar", "Salmonidae")) %>% 
  distinct(X3A_Code, Scientific_Name)

quantity <- quantity %>% 
  merge(species_codes, by.x = "species_alpha_3_code", by.y = "X3A_Code") %>% 
  filter(!is.na(Scientific_Name)) %>% 
  select(-species_alpha_3_code)

# Apply status codes and filter by relevant species groups
status_codes <- file.path(rawdata_path, "Aquaculture_2025.1.0", "CL_FI_SYMBOL_SDMX.csv") %>% 
  read.csv() %>% 
  distinct(Symbol, Name_En, Description_En)

quantity <- quantity %>% 
  merge(status_codes, by.x = "status", by.y = "Symbol") %>% 
  rename(status_nm = Name_En,
         status_description = Description_En) %>% 
  select(-status)
```

```{r rich-raw-processing}
#| eval: false

@@

(mariculture_spp_prod <- aquaculture_raw %>% 
    clean_names() %>% 
    select(c(1:7, starts_with("x"))) %>% 
    pivot_longer(names_to = "year", values_to = "production", cols= -c(1:7)) %>% 
    mutate(year = gsub(pattern="x", replacement = "", year)) %>% 
    filter(year %in% c(2017)) %>% 
    filter(!grepl("Inland waters", fao_major_fishing_area_name)  & 
             !asfis_species_name_3 %in% c("Clams, cockles, arkshells", 
                                          "Abalones, winkles, conchs",
                                          "Mussels", 
                                          "Oysters",
                                          "Shrimps, prawns",
                                          "Crabs, sea-spiders",
                                          "Freshwater molluscs",
                                          "Squids, cuttlefishes, octopuses",
                                          "Pearls, mother-of-pearl, shells",
                                          "Miscellaneous aquatic plants",
                                          "Scallops, pectens"),
           !grepl("seaweeds" , asfis_species_name_3),
           !grepl("molluscs|cucumber|invertebrates", asfis_species_name_2)) %>% 
    group_by(country_name, asfis_species_name_2, asfis_species_name_3, fao_major_fishing_area_name, fao_major_fishing_area_code, year) %>% 
    summarise(production = sum(production)) %>% 
    ungroup() %>% 
    mutate(asfis_species_name_2 = case_when(asfis_species_name_2 == "Marine fishes nei" & country_name == "Viet Nam" ~ "Groupers nei",
                                            TRUE ~ asfis_species_name_2)) %>% 
    mutate(iso3c = countrycode(sourcevar = country_name, origin = "country.name", destination = "iso3c", warn =TRUE)) %>% 
    group_by(iso3c, country_name, year) %>% 
    nest() %>% 
    mutate(national_production = map(data, ~(sum(.$production)))) %>% 
    unnest(cols = c(data, national_production)) %>% 
    ungroup() %>% 
    filter(asfis_species_name_2 %in% spp_names) %>% 
    mutate(prop_production = production/national_production) %>% 
    filter(!asfis_species_name_2 %in% c("Marine fishes nei")) %>% 
  mutate(model_name = case_when(grepl("rouper", asfis_species_name_2) ~ "general_grouper", 
                                asfis_species_name_2 %in% c("Coho(=Silver) salmon", "Rainbow trout") ~ "general_salmonid",
                                asfis_species_name_2 %in% c("Amberjacks nei") ~ "japanese_amberjack",
                                TRUE ~ gsub(pattern = " ", replacement = "_", tolower(asfis_species_name_2)))) %>% 
  relocate(model_name, .before = asfis_species_name_2) %>% 
  filter(prop_production>0) %>% 
    mutate(farm_name = case_when(model_name %in% c("atlantic_salmon", "general_salmonid") ~ "Salmonidae fish",
                                 TRUE ~ "General marine fish"))
    )
```
